Image projections along an axis

### 2006-02-19 article/Article.tex
complete the article and add timing results
darcs-hash:20060219095448-2fc9d-e9b158b89cf455198e0e4075645c3737a1a405be.gz

\section{Introduction}
ITK already provide a class to project an image along an axe:
\doxygen{AccumulateImageFilter}. However, this class is limited to sum
and mean projections.

@@

The projection can be easily threaded by splitting the required region. The
Only important thing is to take care to not split the region on the projection
dimension. 
The default split done by \verb$ImageSource$ if fully usable without any changes
because it is done on the output region, and that region have a size of 1 on
the projection dimension, so the split will be done on anothe dimesion.

AccumulateImageFilter is not modifiable to implement functor support without
breaking the API.

\section{Implemeting a new projection method}
As an example, here is a filter which implement a square sum projection along a
axis. This example may be useful in case of signed input type.

Everything can be done in a single file, itkSquareSumProjectionImageFilter.h.
\small \begin{verbatim}
#ifndef __itkSquareSumProjectionImageFilter_h
#define __itkSquareSumProjectionImageFilter_h
\end{verbatim} \normalsize
The itkProjectionImageFilter.h header must be included.
\small \begin{verbatim}
#include "itkProjectionImageFilter.h"
#include "itkNumericTraits.h"

@@

namespace Function {
\end{verbatim} \normalsize
Now, create the class which do the real work: receive all the pixel values and
give the result at the end. The class is created in the \verb$itk::Function$ names pace
to avoid putting too much things in itk name space, and is templated over the
type of the input and the output pixel.
One object of this class is created for each thread.
\small \begin{verbatim}
template <class TInputPixel, class TOuputPixel>
class SquareSumAccumulator
{
public:
\end{verbatim} \normalsize
The constructor must take one parameter: the number of pixels to accumulate
during the projection. In that case, this parameter is not useful, so it is not
saved in an attribute of the class.

@@

The destructor is empty - there is nothing to do.
\small \begin{verbatim}
  SquareSumAccumulator( unsigned long size ) {}
  ~SquareSumAccumulator(){}
\end{verbatim} \normalsize
\verb$Init()$ is called before accumulating a new set of pixel
to initialize the accumulator. Here, the sum is initialized to zero.
\small \begin{verbatim}
  void Init()
    {
    m_SquareSum = NumericTraits< TOuputPixel >::Zero;
    }
\end{verbatim} \normalsize
The \verb$operator()$ is called for all pixels. In this method, the square sum is
contructed, pixel after pixel. The method is declared inline for best
performance.
\small \begin{verbatim}
  inline TOuputPixel operator()( const TInputPixel &input )
    {
    m_SquareSum = m_SquareSum + vnl_math_sqr( (TOuputPixel) input );
    }
\end{verbatim} \normalsize
\verb$GetValue()$ is called to set the value of the result image. In that example, the
sum is just returned, but more complex computations can be done here (see the
\verb$MedianProjectionImageFilter$'s code for an example). \verb$ProjectionImageFilter$ cast the
output of this method to the output image's pixel type.
\small \begin{verbatim}
  inline TOuputPixel GetValue()
    {
    return m_SquareSum;
    }
\end{verbatim} \normalsize
The internal attribute which accumulate the square sum.
\small \begin{verbatim}
  TOuputPixel m_SquareSum;
};
} // end namespace Function
\end{verbatim} \normalsize
The class is declared. It inherits from ProjectionImageFilter, and declares the
usual types and methods.
\small \begin{verbatim}
template <class TInputImage, class TOutputImage>
class ITK_EXPORT SquareSumProjectionImageFilter :
    public
    ProjectionImageFilter<TInputImage, TOutputImage,
      Function::SquareSumAccumulator< typename TInputImage::PixelType,
        typename TOutputImage::PixelType > >
{
public:
  typedef SquareSumProjectionImageFilter Self;
  typedef ProjectionImageFilter<TInputImage, TOutputImage, 
    Function::SquareSumAccumulator< typename TInputImage::PixelType,
      typename TOutputImage::PixelType > > Superclass;

@@

A timing test comparing performance on a $371 \times 371 \times 34$
image showed that the results are similar with AccumulateImageFilter
and the equivalent ProjectionImageFilter, and that the new projection
types have similar performance. The results achieved on an Athlon 64 
Processor 2800+ (1802 MHz) with 512Kb cache, 512 Mb of RAM and gcc
4.0.2 are shown in Table~\ref{perf}.

\begin{table}[htbp]
\centering
\begin{tabular}{cc}
\hline
Filter & execution time \\
\hline
\hline
MaximumProjectionImageFilter & 0.652 s \\
MinimumProjectionImageFilter  & 0.614 s\\
SumProjectionImageFilter  & 0.614 s\\
MeanProjectionImageFilter  & 0.619 s\\
MedianProjectionImageFilter  & 0.831 s\\
AccumulateImageFilter, sum  & 0.615 s\\
AccumulateImageFilter, mean  & 0.617 s\\
\hline
\end{tabular}
\caption{Execution time.\label{perf}}
\end{table}

\section{Examples}
Figure~\ref{maximum0}, Figure~\ref{maximum1}, Figure~\ref{maximum},
Figure~\ref{minimum}, Figure~\ref{mean}, Figure~\ref{median},
Figure~\ref{sigma} and Figure~\ref{binary100} show the result
of different projection on different axes. The result of
SumProjectionImageFilter in similar to the result of MeanProjectionImageFilter,
but with a higher range of values, and so is not shown here.

@@

\section{Conclusion}
The new classes give more feature and more flexibility with equivalent
performance than the current available filter, but loss the compatibility with
the previous filter. I propose to mark AccumulateImageFilter as deprecated in
the next release of ITK to incite ITK user to use the new class.

>>>>

\section{Introduction}
ITK already provide a class to project an image along an axis:
\doxygen{AccumulateImageFilter}. However, this class is limited to sum
and mean projections.

@@

The projection can be easily threaded by splitting the required region. The
only important thing is to take care to not split the region on the projected
dimension.
The default split done by \verb$ImageSource$ if fully usable without any changes
because it is done on the output region, and that region have a size of 1 on
the projected dimension, so the split will be done on another dimension.

\verb$AccumulateImageFilter$ is not modifiable to implement functor support without
breaking the API. However, \verb$AccumulateImageFilter$ can be implemented as a
sub-class of \verb$ProjectionImageFilter$ to benefit of the thread support, and to
avoid duplication of code.

\section{Implementing a new projection method}
As an example, here is a filter which implement a square sum projection along a
axis. This example may be useful in case of signed input type.

Everything can be done in a single file, \verb$itkSquareSumProjectionImageFilter.h$.
\small \begin{verbatim}
#ifndef __itkSquareSumProjectionImageFilter_h
#define __itkSquareSumProjectionImageFilter_h
\end{verbatim} \normalsize
The \verb$itkProjectionImageFilter.h$ header must be included.
\small \begin{verbatim}
#include "itkProjectionImageFilter.h"
#include "itkNumericTraits.h"

@@

namespace Function {
\end{verbatim} \normalsize
Now, create the class which do the real work: receive all the pixel values and
give the result at the end. The class is created in the \verb$itk::Function$ name space
to avoid putting too much things in the \verb$itk$ name space, and is templated over the
type of the input and the output pixel.
One object of this class is created for each thread.
\small \begin{verbatim}
template <class TInputPixel, class TOuputPixel>
class SquareSumAccumulator
{
public:
\end{verbatim} \normalsize
The constructor must take one parameter: the number of pixels to accumulate
during the projection. In that case, this parameter is not useful, so it is not
saved in an attribute of the class.

@@

The destructor is empty - there is nothing to do.
\small \begin{verbatim}
  SquareSumAccumulator( unsigned long size ) {}
  ~SquareSumAccumulator(){}
\end{verbatim} \normalsize
\verb$Init()$ is called before accumulating a new set of pixels
to initialize the accumulator. Here, the sum is initialized to zero.
\small \begin{verbatim}
  void Init()
    {
    m_SquareSum = NumericTraits< TOuputPixel >::Zero;
    }
\end{verbatim} \normalsize
The \verb$operator()$ is called for all pixels. In this method, the square sum is
constructed, pixel after pixel. The method is declared \verb$inline$ for best
performance.
\small \begin{verbatim}
  inline TOuputPixel operator()( const TInputPixel &input )
    {
    m_SquareSum = m_SquareSum + vnl_math_sqr( (TOuputPixel) input );
    }
\end{verbatim} \normalsize
\verb$GetValue()$ is called to set the value of the pixel in the output image. In that example, the
sum is just returned, but more complex computations can be done here (see the
\verb$MedianProjectionImageFilter$'s code for an example). \verb$ProjectionImageFilter$ cast the
output of this method to the output image's pixel type.
\small \begin{verbatim}
  inline TOuputPixel GetValue()
    {
    return m_SquareSum;
    }
\end{verbatim} \normalsize
The internal attribute which accumulate the square sum.
\small \begin{verbatim}
  TOuputPixel m_SquareSum;
};
} // end namespace Function
\end{verbatim} \normalsize
The class is declared. It inherits from \verb$ProjectionImageFilter$, and declares the
usual types and methods.
\small \begin{verbatim}
template <class TInputImage, class TOutputImage>
class ITK_EXPORT SquareSumProjectionImageFilter :
    public
    ProjectionImageFilter<TInputImage, TOutputImage,
      Function::SquareSumAccumulator< typename TInputImage::PixelType,
        typename TOutputImage::PixelType > >
{
public:
  typedef SquareSumProjectionImageFilter Self;
  typedef ProjectionImageFilter<TInputImage, TOutputImage, 
    Function::SquareSumAccumulator< typename TInputImage::PixelType,
      typename TOutputImage::PixelType > > Superclass;

@@

A timing test comparing performance on a $371 \times 371 \times 34$
image showed that the new filter outperform the current implementation,
while it should have very similar performance.
This is the result of a problem in ITK 2.4.1, where the iterator with index
is a lot more efficient than the one without index. The new implementation
use the iterator with index for best performance, but really don't need
it, and should get even better performance with the iterator without
index once this problem will be fixed.

 The results achieved on an server with 4 Intel(R) Xeon(TM) MP CPU 3.66GHz
with 1024Kb cache, 24 Gb of RAM, hyperthreading activated, gcc
3.4.3 and running Mandriva Linux are shown in Table~\ref{perf} and
Figure~\ref{results}.

\begin{table}[htbp]
\centering
\begin{tabular}{cc}
\hline
Filter & execution time \\
\hline
\hline
\verb$MaximumProjectionImageFilter$ & 0.0841 s \\
\verb$MinimumProjectionImageFilter$  & 0.0846 s\\
\verb$SumProjectionImageFilter$  & 0.0872 s\\
\verb$MeanProjectionImageFilter$  & 0.0828 s\\
\verb$MedianProjectionImageFilter$  & 0.164 s\\
\verb$SigmaProjectionImageFilter$  & 0.123 s\\
\verb$BinaryProjectionImageFilter$  & 0.0848 s\\
\verb$AccumulateImageFilter$, sum  & 0.465 s\\
\verb$AccumulateImageFilter$, mean  & 0.465 s\\
\hline
\end{tabular}
\caption{Execution time with a single thread.\label{perf}}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics{results}
\caption{Execution time with an increasing number of threads.\label{results}}
\end{figure}

Figure~\ref{results} show a important increase of performance while using several
threads\footnote{It also show that the execution time is not really stable
and that the mean of 50 execution time is not enough to compare the execution
time with precision. This will be discussed in another article.}.

Note that above 4 threads, the execution times are still decreasing. In that case,
hyperthreading give an increase of performance.

\section{Examples}
Figure~\ref{maximum0}, Figure~\ref{maximum1}, Figure~\ref{maximum},
Figure~\ref{minimum}, Figure~\ref{mean}, Figure~\ref{median},
Figure~\ref{sigma} and Figure~\ref{binary100} show the result
of different projection on different axes. The result of
\verb$SumProjectionImageFilter$ in similar to the result of \verb$MeanProjectionImageFilter$,
but with a higher range of values, and so is not shown here.

@@

\section{Conclusion}
The new classes give more feature and more flexibility with equivalent
performance than the current available filter, but loss the compatibility with
the previous filter. I propose to mark \verb$AccumulateImageFilter$ as deprecated in
the next release of ITK to incite ITK user to use the new class.
